### 2017-04-03 paper/sexSelectionSpacing-ver1.tex
Begun removing discussion of fertility from intro and theory

\title{Sex-Selective Abortions, Fertility, and Birth Spacing
% \protect\thanks{%
% I am grateful to Andrew Foster and Darryl Holman for discussions about the method.
% I owe thanks to three anonymous referees, Shelly Lundberg, Daniel Rees, David Ribar, 
% Hendrik Wolff, seminar participants at University of Copenhagen, University of Michigan, 
% University of Washington, University of \AA{}rhus, the Fourth 
% Annual Conference on Population, Reproductive Health, 
% and Economic Development, and the Economic Demography Workshop for helpful suggestions and comments.
% I would also like to thank Nalina Varanasi for research assistance.
% Support from the University of Washington Royalty Research Fund and the 
% Development Research Group of the World Bank is gratefully acknowledged.
% The views and findings expressed here are those of the author and
% should not be attributed to the World Bank or any of its member countries.
% Partial support for this research came from a Eunice Kennedy Shriver National
% Institute of Child Health and Human Development research infrastructure grant,
% 5R24HD042828, to the Center for Studies in Demography and Ecology at the
% University of Washington.
% Prior versions of this paper were presented under the title ``The 
% Determinants of sex-selective Abortions.''
% }
}

@@

\begin{document}
\graphicspath{{graphs/},{figures/}}
\DeclareGraphicsExtensions{.jpg,.pdf,.mps,.png}

@@

% \newpage
\doublespacing

@@

% Demography abstract
\noindent This paper addresses two main questions:
what is the relationship between fertility and sex selection 
and
how does birth spacing interact with the use of sex-selective abortions? 
I introduce a statistical method that incorporates how sex-selective abortions 
affect both the likelihood of a son and spacing between births.
Using India's National Family and Health Surveys,
I show that falling fertility intensifies use of sex selection, leading to use at 
lower parities, and longer spacing after a daughter is born.
Women with 8 or more years of education, both in urban and rural areas, are the main users 
of sex-selective abortions and have the lowest fertility.
Women with less education have substantially higher fertility and do not appear to use 
sex selection.
Predicted lifetime fertility for high-education women declined more than 10\% 
between 1985--1994, when sex selection was legal, and 1995--2006, when sex selection was illegal.
Fertility is now around replacement level.
Abortions per woman increased almost 20\% for urban women and 50\%
for rural women between the two periods, suggesting that making sex selection illegal has 
not reversed its use.
Finally, sex selection appears to be used to ensure one son rather than multiple sons.

@@

% India's male to female ratio increased almost continuously during the last century, 
% mainly because of excess mortality of girls \citep{murthi95,dyson01}.

@@

I address two important questions that the prior literature has been unable
to examine because of the simple method's limitations.
First, what is the relationship between fertility decisions and use of sex selection?
Second, how does birth spacing interact with use of sex-selective abortions?
I introduce a novel method that directly incorporates that sex-selective 
abortions affect both the likelihood of a son being born \emph{and} the duration between 
births.
I use the method to argue that differences in fertility over time and between groups
explain a substantial portion of the changes in the use of sex selection in 
India, and to show how spacing between births play an important role in our understanding 
of fertility and sex selection decisions.

We already know that fertility and son preference are related from the substantial literature
on fertility stopping behavior before sex selection was available.
This literature shows that families are more likely to stop childbearing after the 
birth of a son than after the birth of a daughter
\cite[see, for example,][]{Das1987,Arnold1997,clark00}.%
\footnote{
\citet{filmer09} analyse the relationship between the sex composition of previous
children and subsequent fertility behavior using data from 64 countries.
}
It is also easy to see how declining fertility may increase use of sex selection.
Take a family that wants one son.
If the family is willing to have up to 4 children, the probability of having a son
is more than 94 percent, even without sex selection, and that increases to almost 99 
percent if the family is willing to have up to 6 children.%
\footnote{
The probability of not having a son are 48.8 percent for one child, 23.8 percent for
two children, 11.6 percent for 3 children, 5.7 percent for 4 children, 2.8 percent
for 5 children, and 1.4 percent for 6 children.
}
% If the family is willing to have up to six children, the probability of having a son
% is more than 99 percent, even without sex selection.
If the desire is instead for one son \emph{and} a maximum of two children, there 
is a 24 percent chance that the family will have to resort to sex selection to achieve 
both targets.
% Working in the opposite direction, the stated preference for sons appears to decline with 
% lower fertility \citep{das_gupta97,bhat03}.
Despite these opposite targets, there is little empirical analysis of 
the effects of fertility on sex selection using individual level data 
\citep{park95,Ebenstein2011}.%
\footnote{
\citet{Dharmalingam2014} examine how state level fertility in India
relates to desired family size and son preference over time, but does 
not look at how fertility preferences shape the decision on sex-selective 
abortions.
At country level \citet{Bongaarts2013} shows how sex ratios at births are 
only elevated for countries with lower fertility and
\citet{Bongaarts2015} use national level estimates of the relationship
between sex ratio at birth and fertility as part of their
prediction of the number of missing women past and present.
Simulations suggest that in Korea introduction of sex selection
changed family size little, but did result in abortions
of female fetuses equal to about 5 percent of actual female
births \citep{park95}.
For China allowing a three-child policy has been predicted to 
increase the fertility rate by 35 percent, but also reduce the
number of girls aborted by 56 percent \citep{Ebenstein2011}.
}

% Not only can a decline in desired fertility increase the use of sex selection, 
% increasing the availability of prenatal sex determination also allows a 
% woman to reach a desired number of sons with fewer births.
% To fully understand the use of sex selection we therefore need
% to understand both its use by parity \emph{and} by the fertility decision.
% This is especially important if lower fertility is the driver of the
% increased use of sex selection.

% The use of sex-selective abortions may increase over time for a number of different reasons.
% 1. preferences can change to make people desire boys more than before
% 2. the return to girls can fall further 
% 3. desired fertility can fall (even if nothing else changes)
% 3. the cost of sex-selective abortions can fall, which increases its use even though
% there are no other changes.

@@

Closely related to the overall fertility decisions are decisions on  spacing between births.
In the absence of sex-selective abortions, son preference often leads to a shorter duration 
until the next birth if the previous birth was a daughter  
\citep[see, for example,][]{Das1987,Rahman1993,Pong1994,Haughton1996,Arnold1997}.
The resulting shorter spacing is thought to be associated with worse health outcomes for
the girls \citep{arnold98,Whitworth2002,Rutstein2005,Conde-Agudelo2006}.

@@

% [contribution of paper]
The main contribution of this paper is the introduction and application of a method that 
incorporates the decisions about fertility, birth spacing, and sex-selective abortions.
The method allows us to better understand the relationship between these three decisions.
% The method allows us to better understand the relationship between sex-selective abortions, 
% fertility, and birth spacing.
The empirical model is a competing risk, non-proportional hazard model with two exit 
states: either a boy or a girl is born.
This approach has three major advantages over the simple method.
First and foremost, it models fertility and spacing decisions jointly with the birth 
outcome, which makes it possible to better understand how changing fertility decisions
affects the use of sex selection.
Secondly, by explicitly incorporating censoring of birth spacing, it addresses any 
potential bias from ignoring how sex selection decisions may change with duration from 
the previous birth.
Finally, the method allows us to predict completed fertility, total number of abortions, 
and final sex ratio over a woman's reproductive life cycle.
These provide a better way of measuring changes in the use of sex selection 
than what we can get from individual parities---especially because individual 
parities may show a decline in sex selection at the same time as overall
use is increasing.
None of these is possible using the simple method.

% [EXPLAIN RESULTS HERE]
% Main findings!
I apply the method to birth histories from India's National Family and Health Surveys 
for Hindu women covering the period 1972 to 2006. 
There are three main findings.
First, fertility is an important factor in the decision to use sex-selective 
abortions.
On the one hand, women with fewer than 8 years of education have relatively high fertility 
and do not appear to use sex selection.
On the other hand, better-educated women in both urban and rural areas have much lower
fertility and are the main users of sex-selective abortions.
For women with 8 or more years of education and only daughters the 
use is substantial: around 60 percent of the children born are boys.
Furthermore, as better-educated women's fertility has declined over time, sex selection 
occurs for earlier and earlier parities and has correspondingly increased in intensity.
There is, however, still no evidence of sex selection being used on the first birth.

@@

Finally, 
the legal steps taken to combat sex-selective abortions have not been able 
to reverse its use.
In 1994, the Central Government passed the Prenatal Diagnostic Techniques (PNDT) Act 
making determining and communicating the sex of a fetus illegal.%
\footnote{
Abortion has been legal in India since 1971 and still is.
The Act is described in detail at \href{http://pndt.gov.in/}{http://pndt.gov.in/}.
The number of convictions has been low.
It took until January 2008 for the first state, Haryana, to reach 5 convictions.
Hence, private clinics operate with little risk of legal action \citep{Sudha1999}.
Maharashtra was the first state to pass a similar law in 1988.
}
In the period 1985-1994 the predicted total number of abortions per 100 women by the 
end of childbearing was 8 for urban women and 6.8 for rural women.
As completed fertility for high-education women fell to now 
around 2 in urban areas and 2.6 in rural areas, 
the number of sex-selective abortions 100 women are predicted to have 
during their childbearing years has increased to 9.4 for urban women and 
10.5 for rural women.
In other words, close to one in ten well-educated women will now have at least 
one abortion motivated by sex selection during their childbearing.
Hence, although it is possible that the PDNT Act slowed the increase in use 
of sex selection, the total predicted number of abortions per woman continued 
to increase after the law was enacted.
The end result is a sex ratio across all children of 53.7\% boys for well-educated 
urban women and 53.4\% for well-educated rural women.

% This is a substantial increase in sex-selective abortions compared to
% the period before the practice was made illegal nationwide in India in 1994.

\section{Theory\label{sec:model}}

The model I present here serves three purposes.
First, it allows me to illustrate how fertility, sex selection, and birth 
spacing interact when changing the cost of children and access to sex selection.
Second, I show that there are cases where parents reconsider their decision to 
use sex selection between births, which underscores why 
using only observed births may provide a biased view of how much
sex selection takes place and what the final sex ratio will be.
Finally, the model informs the empirical strategy presented below.

In the model, parents decide on fertility sequentially,
given the number of boys and girls they already have.
Parents fully control fertility in each period and the only
source of uncertainty is whether they conceive a boy or a girl.
This means that spacing between births arise only from 
the use of sex selection.%
\footnote{
A more realistic approach would have parents choose how 
``intensive'' they want to use contraceptives, where the
intensiveness translates into a stochastic conception process,
so that parents do not directly control the timing of births 
but rather the probability of having a birth in a given period.
The added complexity would, however, not add anything substantial 
to help answer the main question of what the relationship between fertility 
and sex selection is, and would detract from the main source of uncertainty,
which is whether the couple will conceive a male or female fetus.
Assuming that parents fully control fertility is relatively common 
in the economics literature on dynamic models of fertility, such as
\citet{Happel1984}, \citet{wolpin84}, \citet{rosenzweig85}.
The empirical approach does allow for stochastic occurrence of births
through the use of a hazard model.

See \citet{Leung1991} for a model that incorporates both stochastic 
births based on a chosen level of contraception and differential 
preferences for boys and girls.
The model does not, however, allow for abortions and it does
not appear possible to extend the model to incorporate abortions
and still solve it an similar manner.
}
This is obviously a simplification and since the interaction
between son preference and spacing is at the core of this paper,
I first discuss factors that may affect spacing between birth.%
\footnote{
It is possible that parents care directly about the
length of spacing between births, but 
spacing outcomes are more likely the results of parents weighing many
factors against each other.
}

@@

With the introduction of sex-selective abortions, ``too long'' spacing 
may, however, become a greater concern for parents, as 
repeated abortions can result in very long durations between births.%
\footnote{
Shorter spacing was probably less of a concern to begin with for
better educated mothers---who are also the most likely to use 
sex selection---since their children are substantially less likely 
to be negatively affected by short spacing than children of less 
educated mothers \citep{Whitworth2002}.
In general, mortality risk is negatively related to maternal
education \citep[for India, see, for example,][]{Tulasidhar1993},
although the underlying mechanism is subject
to debate \citep{Kovsted2002}.
}
If ``too long'' spacing leads parents to reverse their decision to use
prenatal sex determination within a spell this has important implications 
for the empirical method.
A reversal may happen, for example, if parents have concerns about possible 
infertility from repeated abortions.
Furthermore, with the increasing education and wealth in India, a growing 
portion of the population is likely to behave in a manner that makes the 
literature on shorter spacing in middle- and high-income countries more relevant.%
\footnote{
For the first birth, a potentially important factor in India is 
the pressure to show fecundity, which would tend to shorten
the duration to the first birth \citep{dyson83,Sethuraman2007,Dommaraju2009}.
}
There may, for example, be economies of scale in childrearing---both
in terms of time cost and direct cost---that are easier to
take advantage of if births are closer \citep[p 947]{Newman1984}.
Similarly, if women's wages increase with age parents prefer
to have their children early and close together while 
the opportunity cost of children is lower \citep{Heckman1976}.%
\footnote{
The time needed to care for young children can also impact 
timing if skills depreciate when out of the labor market \citep{Happel1984}.
Whether depreciation leads to longer or shorter spacing depends
on the specification of the depreciation function.
See also the discussion in \citet[p 315]{Hotz1997}
and references therein.
}
% %
% \footnote{
% As \citet{Heckman1976} show an increasing
% income stream can also lead to longer spacing.
% If children are costly in each period, as in the
% model presented here, parent may use contraception to wait for
% sufficiently higher income before having the next child.
% Notice, that we would expect both the income and the
% cost of children to increase as women's education goes up and
% it is likely that the cost of children increases faster
% than income.
% }

@@

In the absence of prenatal sex determination, stronger son preference is often thought 
to increase overall fertility.
That is not the case here.
In fact, there is little correlation between the weight placed on utility from boys and 
fertility.%
\footnote{
This is confirmed by regressing the weight places on sons, cost of children,
and discount factor against fertility for the expanded set of values.
}
Similarly, parents' discount factor has no impact on fertility in the absence of
prenatal sex determination.
The one factor that does affect fertility is the cost of children;
increasing the cost of children substantially reduces fertility.

@@

Introducing prenatal sex determination has two opposing effects on fertility 
in the model.
On one hand, it takes fewer births to reach a desired number of sons and this 
reduces fertility.
On the other hand, parents no longer have to pay for unwanted children and 
this ``wealth effect'' increases fertility.
% The simulations show a small negative effect of lowering the cost of
% sex selection holding the other parameters constant.
The simulations show that for most combinations of son preference and 
cost of children cutting the cost of abortion has little effect 
on fertility---even if the cut is substantial---indicating that for this 
set-up and the chosen parameter values the two effects 
cancel each other out.%
\footnote{
Running regressions on the expanded simulation results show a negative, but
statistically insignificant relationship between cost of abortion
and fertility.
For high son preference and cost of children the effect of 
decreasing cost of prenatal scan is either nill or slightly
positive.
See also the discussions in \cite{leung94} and \cite{davies97}.
}

@@

% Cost of children

The strongest driver of fertility in the model is the cost of
child maintenance per period.
% For a period maintenance cost of 12 fertility is between 7 and 7.5,
% while for a cost of 22 fertility it is only 2.
As fertility decreases with higher cost of children, sex ratios 
become more male biased---although it is possible that there is 
non-linearity in the effect of child cost on sex ratios.%
\footnote{
The changes in sex ratio are not uniformly positive, possibly because 
the discrete nature of fertility decisions combined with the uncertainty 
of the sex of the next conception produce knife-edge cases that can
dramatically swing the outcomes.
}
This non-linearity also shows up in the number of abortions 
and spells where a prenatal scan has been used, 
but the overall trend is towards more abortions and more spells with
prenatal scan per birth as the cost of children increase.

@@

% The non-uniform effect in spacing by parity when changing
% the cost of prenatal scan show up here as well.
% Consider, for example, the case when cost of prenatal scan falls from 20
% to 16 when cost of children is 16. 
% The average spacing for parity 5 births falls from 1.44 to 1.18, but
% the spacing for parity 4 births increases only from 1 to 1.09. 
% What this hides is, of course, that there are substantially more 
% parity 4 births than there are parity 5 births--and only a slight 
% decline in fertility from 4.16 to 4.11---and that the number of
% abortions per 100 woman doubles.

% non-uniform effect in spacing by parity
% Interestingly, something similar happens when there is no son preference
% See a=0.5, k = 18 all p. Fourth parity reduces throughout, but third parity
% increases throughout.

% from p= 20 to p=16 it looks like sex selection is falling overall
% because the average spacing goes down substantially for parity 5 (1.44 to 1.18)
% and spacing for parity 4 only increases from 1 to 1.09

% Less clear but generally similar is k = 14]

% The problem with testing this idea is that balance would, for example,
% show up as people using sex selection on having 2 girls and 1 boy and
% the desired fertility is 4. 
% But, if people have high alpha then we would get the same outcome.

% [something on the importance of this and the idea that are 
% multiple forms of son preference.
% people may treat their children equally, but would not abort
% a male to ensure equal number of sons and daughters]

% Discount rate

The discount rate affects fertility little, but sex ratios 
become more biased the more patient people are (higher $d$).%
\footnote{
If son preference is very strong the effect appear to 
depend on the cost of children. 
For low cost of children higher impatience leads to higher
fertility, but for higher cost of children it is sometimes reversed.
Furthermore, these effects are not very strong or clear and
potentially non-linear.
}
% The motivation for this is that less weight you place on the future
% the more appealing it is to simply have another child instead of
% paying for prenatal sex determination, which is ``wasted'' if
% you have a boy.
Similarly, the number of abortions and the number of spells where
prenatal scans are used are higher the more 
weight is place on future utility over current utility.
Just like the discount rate, increasing parents'  
horizon (larger $T$) changes fertility little, but does tend to 
increase the sex ratio in favor of boys when there is access 
to sex selection.%
\footnote{
These simulation results are presented in the online Appendix.
}
Correspondingly, the number of abortions are higher the 
longer the time horizon.

@@

% What are the model highlights 
There are three main conclusions from the model.%
\footnote{
A caveat is that these results reflect the particular utility 
function chosen.
If parents have a different utility function, like a preference
for one son but otherwise no preference across boys and girls
the results may look different.
% As shown it still is possible, however, to get sex selection
% for balance and for parents to change their decision on use
% of sex selection between births.
}
First, the strongest driver of increased use of sex selection is 
the fertility decline that arise from an higher cost of 
children---for example because of more education for women or 
migration to urban areas.
% As discussed, an important predictor of cost of children is the
% mother's education---more educated women have higher opportunity 
% cost of time.
The increased use of sex selection shows up both in the 
proportion of boys among births and longer spacing between 
births---and the higher the parity the longer the duration between 
births.
The other factors, such as the price of sex selection, 
discount factor, and son preference also affect the use
of sex selection, but to a much smaller degree.
Second, parents do reverse the use of sex selection, 
both when they use sex selection relatively infrequently and when 
they use it heavily.
Reversal are more likely the more impatient parents are or the
shorter time they have left in their reproductive horizon, which
means that the longer the duration has been since the last birth
the more likely they are to reverse their decision to use sex selection.
% Furthermore, as the simulations show it is possible for son 
% preference to decline while the sex ratios becoming more unequal 
% provided the cost of prenatal scan decreases and/or cost of children 
% increases sufficiently (or people become more patient).
Finally, the model shows that son preference in the number of
children is not required for an unequal sex ratio.
As long as parents are only willing to abort female fetuses, declining
fertility will lead to increased use of sex selection to ensure 
a balanced number of sons and daughters.

% Wonder if it would be possible to that some of the
% couples with a son already that look like they are
% using sex selection were not planning on having the
% last child, but with prenatal testing and access to
% abortion they would abort a female fetus but not
% a male one, even if the pregnancy was ``unintended''
% (as in having selected a high $e$, but random 
% chance lead to a pregnancy anyway.

% Empirical method
The model highlights three major limitations of the simple 
empirical method, which a better empirical approach needs to 
allow for.
% First, as fertility declines the proportion of boys among births 
% increases and the spells between births become longer.
% Second, the higher the parity the longer the duration between births because of
% the more intensive use of sex selection.
% Finally, households may reverse its decision to use sex selection within a spell.
First, the simple model cannot capture changes in fertility and the extent 
to which these fertility changes drive the use of sex selection.
Second, increased use of sex selection leads to longer spacing which,
everything else equal, reduces the number of births available for 
estimation in the simple model and thereby lowers its precision
because fewer parents will make it to a given parity by the time of the survey.
Furthermore, the more intensively sex selection is used the larger 
the decline in sample size.
Finally, by using only observed births the simple method cannot capture
when parents change their decision on the use of sex selection;
with a reversal the spell can still result in the birth of a girl.
This last point is especially important when trying to establish whether 
son preference has changed over time in a situation where the cost of prenatal 
sex determination has likely \emph{increased} over time because of the PNDT act.
As I show an increase the cost of prenatal sex determination does 
decrease the sex ratio and lower the number of abortions, but 
it also \emph{increases} the likelihood that parents change their initial
decision to use sex selection within a spell.

@@

\subsection{Desired Fertility and Sex Selection}

A first step to understand the relationship between fertility and sex selection
is to examine how use of sex-selective abortions vary with self-reported desired 
fertility.
If lower fertility drives sex selection and desired and actual fertility are
at least partly correlated, women with lower desired fertility should be more 
likely to use sex selection than women with higher desired fertility, 
holding other characteristics constant.
Figure \ref{fig:results_spell2_high_want} presents results by desired fertility 
for the second spell of women with 8 or more years of education whose first child was a 
girl.%
\footnote{
For women with living children the question asked was
``If you could go back to the time you did not have any children and could choose exactly 
the number of children to have in your whole life, how many would that be?'', while
for women with no living children the question was
``If you could choose exactly the number of children to have in your whole life, how many 
would that be?''
% P 116 Appendix F in NFHS book
}

A majority of better-educated women report that they want two or fewer 
children, and the proportion who wants two or fewer is increasing over time.%
\footnote{
In the first period 2,055 out of 8,202 (25\%) women wanted 3 or more 
children, including giving a non-numeric response.
In the second period this number was 2,583 out of 16,600 (16\%) 
and in the third period it was 1,264 out of 13,610 (9\%).
} 
To achieve a large enough sample of women who want three or more children, I use all 
second spells that began in the period 1985 to 2006.
This allows me to estimate non-proportional hazards by area of residence, sex of 
previous child, and desired fertility.
For both rural and urban women, the left panel shows the results for women who want two 
or fewer children and the right panel shows the results for women who 
want three or more children.%
\footnote{
Restricting to wanting exactly 3 children has little to no impact on the spacing
pattern, suggesting that higher desired number of children does not lead to
substantially shorter spacing.
}

>>>>

\title{Sex-Selective Abortions and Birth Spacing
\protect\thanks{%
I am grateful to Andrew Foster and Darryl Holman for discussions about the method.
I owe thanks to three anonymous referees, Shelly Lundberg, Daniel Rees, David Ribar, 
Hendrik Wolff, seminar participants at University of Copenhagen, University of Michigan, 
University of Washington, University of \AA{}rhus, the Fourth 
Annual Conference on Population, Reproductive Health, 
and Economic Development, and the Economic Demography Workshop for helpful suggestions and comments.
I would also like to thank Nalina Varanasi for research assistance.
Support from the University of Washington Royalty Research Fund and the 
Development Research Group of the World Bank is gratefully acknowledged.
The views and findings expressed here are those of the author and
should not be attributed to the World Bank or any of its member countries.
Partial support for this research came from a Eunice Kennedy Shriver National
Institute of Child Health and Human Development research infrastructure grant,
5R24HD042828, to the Center for Studies in Demography and Ecology at the
University of Washington.
% Prior versions of this paper were presented under the title ``The 
% Determinants of sex-selective Abortions.''
}
}

@@

\begin{document}
\graphicspath{{figures/}}
\DeclareGraphicsExtensions{.jpg,.pdf,.mps,.png}

@@

\newpage
\doublespacing

@@

% Demography abstract
\noindent This paper addresses the question:
how does birth spacing interact with the use of sex-selective abortions? 
I introduce a statistical method that incorporates how sex-selective abortions 
affect both the likelihood of a son and spacing between births.
Using India's National Family and Health Surveys,
I show that use of sex selection leads to longer spacing after a daughter is born.
Women with 8 or more years of education, both in urban and rural areas, are the main users 
of sex-selective abortions.
Women with less education do not appear to use sex selection.
Predicted lifetime fertility for high-education women declined more than 10\% 
between 1985--1994, when sex selection was legal, and 1995--2006, when sex selection was illegal.
Abortions per woman increased almost 20\% for urban women and 50\%
for rural women between the two periods, suggesting that making sex selection illegal has 
not reversed its use.
Finally, sex selection appears to be used to ensure one son rather than multiple sons.

@@

@@

I address an important questions that the prior literature has been unable
to examine because of the simple method's limitations:
how does birth spacing interact with use of sex-selective abortions?
I introduce a novel method that directly incorporates that sex-selective 
abortions affect both the likelihood of a son being born \emph{and} the duration between 
births.
I use the method to show how spacing between births play an important role in our 
understanding of sex selection decisions.

@@

In the absence of sex-selective abortions, son preference often leads to a shorter duration 
until the next birth if the previous birth was a daughter  
\citep[see, for example,][]{Das1987,Rahman1993,Pong1994,Haughton1996,Arnold1997}.
The resulting shorter spacing is thought to be associated with worse health outcomes for
the girls \citep{arnold98,Whitworth2002,Rutstein2005,Conde-Agudelo2006}.

@@

% [contribution of paper]
[TK maybe not focus so much on the simple method but rather understanding son preference].
The main contribution of this paper is the introduction and application of a method that 
incorporates the decisions about birth spacing, and sex-selective abortions.
The empirical model is a competing risk, non-proportional hazard model with two exit 
states: either a boy or a girl is born.
This approach has TK major advantages over the simple method

Secondly, by explicitly incorporating censoring of birth spacing, it addresses any 
potential bias from ignoring how sex selection decisions may change with duration from 
the previous birth.

None of these is possible using the simple method.

% [EXPLAIN RESULTS HERE]
% Main findings!
I apply the method to birth histories from India's National Family and Health Surveys 
for Hindu women covering the period 1972 to 2006. 
There are three main findings.

There is, however, still no evidence of sex selection being used on the first birth.

@@

\section{Theoretical Framework\label{sec:model}}

[TK Calling this a model will be overkill - save the model for the fertility paper]

[TK find examples of the son preference spacing papers and look at what they did
for "theory"]

The model I present here serves two purposes.
First, the model informs the empirical strategy presented below.
Second, I show that there are cases where parents reconsider their decision to 
use sex selection between births, which underscores why 
using only observed births may provide a biased view of how much
sex selection takes place and what the final sex ratio will be.

[move discussion of spacing from footnote above to here?]

This is obviously a simplification and since the interaction
between son preference and spacing is at the core of this paper,
I first discuss factors that may affect spacing between birth.%
\footnote{
It is possible that parents care directly about the
length of spacing between births, but 
spacing outcomes are more likely the results of parents weighing many
factors against each other.
}

@@

With the introduction of sex-selective abortions, ``too long'' spacing 
may, however, become a greater concern for parents, as 
repeated abortions can result in very long durations between births.%
\footnote{
Shorter spacing was probably less of a concern to begin with for
better educated mothers---who are also the most likely to use 
sex selection---since their children are substantially less likely 
to be negatively affected by short spacing than children of less 
educated mothers \citep{Whitworth2002}.
In general, mortality risk is negatively related to maternal
education \citep[for India, see, for example,][]{Tulasidhar1993},
although the underlying mechanism is subject
to debate \citep{Kovsted2002}.
}
If ``too long'' spacing leads parents to reverse their decision to use
prenatal sex determination within a spell this has important implications 
for the empirical method.
A reversal may happen, for example, if parents have concerns about possible 
infertility from repeated abortions.
Furthermore, with the increasing education and wealth in India, a growing 
portion of the population is likely to behave in a manner that makes the 
literature on shorter spacing in middle- and high-income countries more relevant.%
\footnote{
For the first birth, a potentially important factor in India is 
the pressure to show fecundity, which would tend to shorten
the duration to the first birth \citep{dyson83,Sethuraman2007,Dommaraju2009}.
}
There may, for example, be economies of scale in childrearing---both
in terms of time cost and direct cost---that are easier to
take advantage of if births are closer \citep[p 947]{Newman1984}.
Similarly, if women's wages increase with age parents prefer
to have their children early and close together while 
the opportunity cost of children is lower \citep{Heckman1976}.%
\footnote{
The time needed to care for young children can also impact 
timing if skills depreciate when out of the labor market \citep{Happel1984}.
Whether depreciation leads to longer or shorter spacing depends
on the specification of the depreciation function.
See also the discussion in \citet[p 315]{Hotz1997}
and references therein.
}

@@

@@

@@

@@

@@

% What are the model highlights 
There are three main conclusions from the model.%
\footnote{
A caveat is that these results reflect the particular utility 
function chosen.
If parents have a different utility function, like a preference
for one son but otherwise no preference across boys and girls
the results may look different.
% As shown it still is possible, however, to get sex selection
% for balance and for parents to change their decision on use
% of sex selection between births.
}
The increased use of sex selection shows up both in the 
proportion of boys among births and longer spacing between 
births---and the higher the parity the longer the duration between 
births.
The other factors, such as the price of sex selection, 
discount factor, and son preference also affect the use
of sex selection, but to a much smaller degree.
Second, parents do reverse the use of sex selection, 
both when they use sex selection relatively infrequently and when 
they use it heavily.
Reversal are more likely the more impatient parents are or the
shorter time they have left in their reproductive horizon, which
means that the longer the duration has been since the last birth
the more likely they are to reverse their decision to use sex selection.
% Furthermore, as the simulations show it is possible for son 
% preference to decline while the sex ratios becoming more unequal 
% provided the cost of prenatal scan decreases and/or cost of children 
% increases sufficiently (or people become more patient).
Finally, the model shows that son preference in the number of
children is not required for an unequal sex ratio.
As long as parents are only willing to abort female fetuses, declining
fertility will lead to increased use of sex selection to ensure 
a balanced number of sons and daughters.

% Empirical method
The model highlights two major limitations of the simple 
empirical method, which a better empirical approach needs to 
allow for.
First, increased use of sex selection leads to longer spacing which,
everything else equal, reduces the number of births available for 
estimation in the simple model and thereby lowers its precision
because fewer parents will make it to a given parity by the time of the survey.
Furthermore, the more intensively sex selection is used the larger 
the decline in sample size.
Second, by using only observed births the simple method cannot capture
when parents change their decision on the use of sex selection;
with a reversal the spell can still result in the birth of a girl.
This last point is especially important when trying to establish whether 
son preference has changed over time in a situation where the cost of prenatal 
sex determination has likely \emph{increased} over time because of the PNDT act.
As I show an increase the cost of prenatal sex determination does 
decrease the sex ratio and lower the number of abortions, but 
it also \emph{increases} the likelihood that parents change their initial
decision to use sex selection within a spell.

@@

\subsection{Comparing the Hazard and Simple Models}

As detailed in the main paper, one of the main advantages of the hazard model 
approach is that it can be used to predict what completed fertility, number of abortions, 
and sex ratios will be once women are done with childbearing.
This cannot be done with the simple model because it does not predict fertility
progression, does not take into account censoring, and cannot capture if parents
change their use of sex selection within a spell.
% This is possible because the hazard model approach uses both censored and 
% uncensored observations, which, in turn, allows me to predict both how many 
% women will have a next birth and the final sex ratio for that parity, under the 
% assumption that the patterns of spacing and use of sex selection remain 
% the same within a spell.
We can, however, still compare differences in predicted sex ratios for 
\emph{individual} parities.%
\footnote{
The simple model uses only observed births and the predicted sex ratio is 
therefore simply what we observe for the relevant uncensored sample.
}

For each parity, the simple model is
\begin{equation}
Y_i = \gamma + \alpha' \mathbf{Z_i} + \beta' \mathbf{X}_{i} + \epsilon_i,
\end{equation}
where $Y$ is a dummy variable that takes the value 1 if the child born is a boy and zero 
if it is a girl, and the vectors of explanatory variables, $\mathbf{Z}_{i}$ and 
$\mathbf{X}_{i}$, are the same as for the proposed method, except that 
$\mathbf{Z}_{i}$ obviously does not include the baseline hazard and that there are 
no time dependence.%
\footnote{
To ease presentation, the indicator for parity number is not shown.
}
I estimate this using a logit model, although a linear probability model leads to
similar results.
This model is estimated on all observed births for a given parity in the relevant 
period to ensure that the results are as close to the standard method of 
examining sex selection \citep[see, for example, ][]{retherford03b,jha06,abrevaya09}.

% Two factors determine the differences in the predicted sex ratio and number 
% of abortions between the two models.
% First, how many women have censored spells but will have a birth after the survey.
% Second, how much use of sex selection changes within a spell.
% % If there are few censored spells or the women with the censored spells have low likelihoods
% % of having a next birth the two models produce similar predicted sex ratios and 
% % number of abortions.
% % If there is little change in the use of sex-selective abortions over a spell then the
% % predicted sex ratio is going to be similar for the two models, although the predicted
% % number of abortions will differ depending on how many censored observations there are
% % and how likely they are to exit with a birth.
% Consider four possible combinations of the two factors.
% First, there are few censored birth spells and use of sex selection is constant over
% the spell.
% In that case, the predictions will be similar across the two models.
% Second, there are many censored birth spells and use of sex selection is constant over
% the spell.
% The predicted sex ratios will be similar, but the simple model will predict a too low
% number of abortions per woman and fertility will appear to be falling faster than it is.
% Third, there are few censored birth spells and use of sex selection changes substantially
% with spell length.
% The sex ratios again are close to each other in the two models, but the estimated use of
% abortions will be biased in the simple model because it cannot account for the changes
% in use of sex selection over the spell.
% Finally, there are many censored birth spells and use of sex selection changes substantially
% with spell length.
% In this case the predicted sex ratio based on the simple model will be biased.
% Specifically, if use of sex selection declines over the spell, the simple model 
% will predict a higher final sex ratio than the hazard model.
% Similarly, the predicted number of abortions will also be biased. 
% With sex selection and falling fertility censored spells become more and more likely,
% but how much this matter is an empirical question.

% To determine the predicted number of abortions I use the natural sex ratio.
% On average, 105 boys should be born per 100 girls born, which corresponds exactly
% with what we observe for first births in India as seen above.
% Hence, if we observe $b$ boys, then we should expect to see $b\frac{100}{105}$ girls.
% If we observe $g$ girls the number of abortions is therefore $b\frac{100}{105}-g$.
% The standard approach is to use the observed sex ratio for a given parity and 
% calculate the number of abortions using this formula.
% For the hazard model it is, however, possible to calculate the number of predicted
% abortions for each time period based on the predicted sex ratio and the 
% number of women predicted to exit with a birth in that period using the same formula.
% Summing across the entire period covered by the hazard model provides a predicted total 
% number of abortions.%
% \footnote{
% There is no difference between the two method in the predicted number of abortions for 
% those women for whom we observe the subsequent birth.
% }
